Lossy compression of DNA sequencing quality data

### 2015-02-03 manuscript/q-compression.tex
More revisions describing the approaches we explore to emphasize the generalize the concepts. The approaches are not called binning, modeling, and profiling.

We examine three general classes of lossy compression methods:
binning, model-based, and profile-based.

@@

Because we only have two possible choices for a quality of base, this
makes a straightforward binary encoding possible, allowing us to use a
single bit to represent the quality of a base instead of the default 8
bits used to store quality values in \textsc{ascii}.

@@

\subsection{Polynomial regression}

One approach to compressing quality values is to treat them as
bivariate data with the ordered nucleotides (1 to read length)
representing the abscissa, and quality values representing the
ordinate. Viewed in this way we can ask the question: is there a
mathematical function that describes the relationship between read
position and quality value sufficiently, albeit imperfectly (i.e.,
with some loss of information), such that modeled values perform well
in subsequent analyses?

To this end, we modeled read quality sequences with zero to
fifth-degree polynomials obtained with least-squares fitting. However,
despite the fact that polynomial functions have significantly lower
number of parameters (i.e. one to six coefficients) than a read-length
string of raw quality values, the necessity of using floating point
numbers to store coefficients greatly limits the compression potential
of the method. In order to get meaningful compression on
single-precision four-byte floating point numbers, one would have to
relax on the least-squares approximation constraint to obtain
compressible values on the byte level which is outside the scope of
this study.

Nonetheless, fitting a polynomial and reconstructing the quality
values that it would have yielded after decompression still makes some
sense as it alters the properties of the quality data, making it
highly compressible simply due to the smoothing effect it
provides. Therefore the compression ratio and bits/base-call values
for polynomial regression throughout this manuscript refer to
polynomially approximated quality values compressed with zip.

\subsection{Profile-based compression}

@@

\subsection{Compressibility}

>>>>

We examine three general classes of lossy compression methods:
binning, modeling, and profiling.

@@

Because we only have two possible choices for a quality of base, this
makes a straightforward binary encoding possible, allowing us to use a
single bit to represent the quality of a base instead of the default 8
bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.

@@

\subsection{Modeling}

If quality values are modeled compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. In this
research we model read quality values as polynomial functions obtained
with least-squares fitting, as one approach to compression read
quality values by modeling.

Despite the fact that polynomial functions have significantly lower
number of parameters (i.e. one to six coefficients) than a read-length
string of raw quality values, the necessity of using floating point
numbers to store coefficients greatly limits the compression potential
of the method. In order to get meaningful compression on
single-precision four-byte floating point numbers, one would have to
relax on the least-squares approximation constraint to obtain
compressible values on the byte level which is outside the scope of
this study.

\subsection{Profiling}

@@

\subsection{Compressibility versus error}
